GEOG0149 - Urban Policy Practical

Is there a relationship between high levels of homeownership and unemployment across Europe?

In recent decades, governments across the Global North have used a range of policy levers to try and increase homeownership rates. This thrust has been based on a belief that homeownership improves prosperity and wealth accumulation, fosters good citizenship, helps build more stable communities, and is simply the tenure of preference for most people.

While there is much debate about the extent to which homeownership actually provides these personal and collective benefits, there has been surprisingly little consideration of the adverse impact that promoting owner-occupation could have on the labour market. This is in part because research on the economics of tenure dynamics has concentrated heavily on housing wealth and its role in welfare provision.

There are good reasons to think that high rates of homeownership could adversely affect labour market performance. In a series of influential papers, Andrew Oswald showed that higher homeownership in an area or country is associated with a higher unemployment rate. He posited several mechanisms that could explain this pattern (we covered these in lecture 9 – they include the lower residential mobility of homeowners reducing job search and job matching; longer commutes by homeowners depressing productivity; reduced business formation in homeownership dominated areas etc) and provided correlational evidence for a positive association between homeownership and unemployment at the level of American states and OECD countries. Other studies have corroborated these findings, with Borg and Branden (2018) arguing that the smaller size of labour markets with high homeownership rates reduces the efficiency of job matching, raising the probability of unemployment for residents regardless of their tenure (an externality effect). Interestingly though, at the micro-level research consistently finds that homeowners are less likely to be unemployed than tenants. So what’s going on?

Today you are going to take a fresh look at this puzzle using recent data on European countries. We will begin by updating Oswald’s work to see if there is still an association between higher homeownership and higher unemployment rates in the aftermath of the Global Economic Crisis. We will then examine change over time in both variables to try and get a better understanding of how and why homeownership and unemployment might be interlinked. Our overarching aim is to try and replicate Oswald and Blanchflower’s approach using fresh data.

1. Load Libraries

```{r}
# load the 'here' package for file path management
library(here)
# load the 'tidyverse' package for data manipulation, exploration, and visualisation
library(tidyverse)
# load the 'stargazer' package to create well-formatted tables
library(stargazer)
# load the 'ggrepel' library to address label overlap
library(ggrepel)
```

2. Load Datasets

```{r}
# read in Unemployment_Europe.csv
unemployment <- read_csv(here("Unemployment_Europe.csv"))
# read in Ownership_Europe.csv
ownership <- read_csv(here("Ownership_Europe.csv"))
```

3. Merge Datasets

```{r}
# merge the 'unemployment' and 'ownership' dataframes by the common column "Country"
europe_rates <- merge(unemployment, ownership, by="Country")
```

4. Initial Exploration

```{r}
# perform a Pearson correlation test between the 2011 unemployment rate and home ownership rate
cor.test(europe_rates$Unemp_2011,
         europe_rates$Own_2011,
         method = "pearson")

# correlation coefficient of 0.452 with a p-value of 0.012 suggests a statistically significant, moderate positive relationsships between unemployment and home ownership rates in 2011 across European countries
# the p-value (0.012) is statistically significant at the 5% level (p < 0.05), meaning we can reject the null hypothesis that there is no correlation
```

```{r}
# ensure reproducibility by fixing the random seed. This helps geom_text_repel() place labels consistently across plots
set.seed(42)

# visualise the relationship between homeownership and unemployment rates in 2011 across European conuntries
ggplot(data=europe_rates, aes(Own_2011, Unemp_2011)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = 0.7) +
  geom_text_repel(aes(label=Country), size = 3.5, max.overlaps = 100) +
  labs(x = "Homeownership Rate in 2011 (%)", y = "Unemployment Rate in 2011 (%)") +
  theme_classic() +
  theme(axis.text = element_text(size = 8))
  
# the regression line indicates a moderate positive correlation: countries with higher homeownership rates tend to have higher unemployment rates
# the shaded area represents the 95% confidence interval around the regression line
# Spain, Greece, Lithuania, and Latvia show high unemployment & high homeownership
# Switzerland, Austria, and Germany show low unemployment and low homeownership
# Romania and Norway are outliers in that they have very high homeownerhsip but relative low unemployment
# the plot supports the hypotheis that high homeownership may be associated with reduced labour mobility, which in turn may contribute to higher unemployment, as people are less likely to move for work
# outliers like Norway or Romania suggest the relationship is not uniform and may be influenced by other factors
```

```{r}
# transform the dataset from wide to long format
europe_rates_long <- pivot_longer(europe_rates, cols = Unemp_2011:Own_2018, names_to = c(".value", "Year"), names_sep = "_")
```

```{r}
# calculate correlations by year
# n() returns the number of observations for each year
group_by(europe_rates_long, Year) %>%
  summarise(cor = cor(Unemp, Own), n = n()) %>%
  print()
```

```{r}
# plot unemployment vs homeownership by year
ggplot(europe_rates_long, aes(Own, Unemp)) +
  geom_point(colour = "navy", size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = 0.7) +
  labs(x = "Homeownership Rate (%)", y = "Unemployment Rate (%)") +
  theme_classic() +
  theme(axis.text = element_text(size = 8)) +
  facet_wrap(~Year)

# between 2011-2014, the relationship is clearly positive, supporting the hypothesis that high homeownership may hinder labour mobility
# in 2015 and 2016, the positive association starts to weaken, possiby due to labour market recovery or housing reforms in some countries
# in 2017 and 2018, the relationship becomes very weak or even flat, suggesting no clear association between homeownership and unemployment in these years
# the positive relationship observed in early years (especially post-2008 financial crisis) dissipates over the period, possibly due to more flexible employment models, urban housing reforms, or policy interventions promoting labour mobility
```

```{r}
# calculate the change in unemployment and change in homeownership between 2011 and 2018
# the lage of 7 rows corresponds to a 7-year difference from 2011 to 2018
# keep only the 2018, so we're left with the change from 2011 to 2018 per country
changes <- europe_rates_long %>%
  group_by(Country) %>%
  mutate(Change_unemp = Unemp - lag(Unemp, n =7), Change_own = Own - lag(Own, n = 7)) %>% 
  filter(Year == 2018) %>% 
  select(Country, Change_unemp, Change_own)

# inspect
changes
```

```{r}
# set seed
set.seed(42)

# vissualise how change in homeownership rates relate to changes in unemployment rates across European countries from 2011 to 2018
ggplot(data=changes, aes(Change_own, Change_unemp)) +
  geom_point() +
  geom_text_repel(aes(label=Country), size = 3.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = 0.7) +
  labs(x = "Change in Homeownership Rate in 2011-2018", y = "Change in Unemployment Rate in 2011-2018") +
  theme_classic() +
  theme(axis.text = element_text(size = 8)) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_x_continuous(limits=c(-10, 5)) +
  scale_y_continuous(limits=c(-10, 5))

# Changes in homeownership rates are only slightly positively correlated to changes in unemployment rates from 2011 to 2018
# countries like Ireland, Lithuania, and Spain show falling unemployment regardless of whether homeownership increased or decreased
# many countries have seen both falling homeownership rates and falling unemployment rates
# Denmark had a notable drop in homeownership and only a modest fall in unemployment
# Ireland experienced the largest drop in unemployment with almost no change in homeownership
# Luxembourg saw a sharp rise in homeownership with little changes in unemployment
```

```{r}
# fit a linear regression model to examine the relationship between changes in homeownership and changes in unemployment from 2011 to 2018
linear_1 <- lm(Change_unemp ~ Change_own, data = changes)

# display the summary of the regression model, including coefficients, R-squared value, and statistical significance
summary(linear_1)

# the summary results show that the relationship weakened consistenyl between 2011 and 2018
# on average, countries saw a decline of 3.12 percentage points in unemployment between 2011-2018 when there wass no change in homeownerhsip
# slope: 0.16 (not significant, p = 0.56) - the relationship between changes in homeownership and changes in unemployment is weak and not statistically significant
# R-Squared of 0.01 shows that only 1% of the variation in unemployment change is explained by homeownership change. This is very weak explanatory power.
```

5. Bring in External Explanatory Variables

```{r}
# read in Age65plus_Europe.csv
age <- read_csv(here("Age65plus_Europe.csv"))
# read in Education_Europe.csv
education <- read_csv(here("Education_Europe.csv"))
```

```{r}
# transform the dataset from wide to long format
age_long <- pivot_longer(age, cols = Age65plus_2011:Age65plus_2018, names_to = "Year", names_prefix = "Age65plus_", values_to = "Age65plus")

# replace 'Slovak Republic' with 'Slovakia' to be consistent with other datasets
age_long <- age_long %>% 
  mutate(Country = ifelse(Country == "Slovak Republic", "Slovakia", Country))

# inspect
# Liechtenstein is missing from other datasets
# Croatia is missing from education_long dataset
age_long
```


```{r}
# transform the dataset from wide to long format
education_long <- pivot_longer(education, cols = Degree_2011:Degree_2018, names_to = "Year", names_prefix = "Degree_", values_to = "Degree")

# inspect
education_long
```

```{r}
# quck check for mismatched or missing countries
setdiff(age_long$Country, education_long$Country) # Liechtenstein missing from education_long
setdiff(education_long$Country, age_long$Country) # Croatia missing from age_long
```


```{r}
# merge age_long and education_long to our original europe_rates_long dataset
europe_rates_long_merged <- europe_rates_long %>% 
  merge(., age_long, by = c("Country", "Year")) %>% 
  merge(., education_long, by = c("Country", "Year"))
```

```{r}
# inspect - 232 rows remaining out of 240 (Croatia dropped. Liechtenstein was not in the original dataset)
europe_rates_long_merged
```


```{r}
# add a new column Log_unemp to noramlise exponential relationships
europe_rates_long_merged <- mutate(europe_rates_long_merged, Log_unemp = log(Unemp))
```

```{r}
# check that the Log_unemp variable has a more normal distribution than the raw Unemp rate variable
# histogram of raw unemployment rate
ggplot(europe_rates_long_merged, aes(Unemp)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Unemployment Rate", x = "Unemployment Rate (%)", y = "Count") +
  theme_minimal()

# histogram of log-transformed unemployment rate
ggplot(europe_rates_long_merged, aes(Log_unemp)) +
  geom_histogram(binwidth = 0.2, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Log(Unemployment Rate)", x = "Log(Unemployment Rate)", y = "Count") +
  theme_minimal()
```

```{r}
# compute the lagged homeownership and unemployment rates for use in the model
# These variables just measure (i) homeownership and (ii) unemployment in the country in the previous year
europe_rates_long_merged <- europe_rates_long_merged %>%
  group_by(Country) %>%
  mutate(Lag_own = lag(Own), Lag_log_unemp = lag(Log_unemp))
```

```{r}
# fit a linear model predicting log unemployment as a function just of lagged homeownership
linear_2 <- lm(Log_unemp ~ Lag_own, data = europe_rates_long_merged)

# use stargazer to present results
stargazer(linear_2, type = "text")

# for every 1 percentage point increase in lagged homeownership, the logged unemployment rate increae by 0.009 on average, holding everything else constant
# since log-unemployment is the outcome, this effect translate to a 0.9% increase in the unemployment rate per 1% increase in lagged homeownership
# the result is statistically significant at the 1% level
# at the baseline (when Lag_own = 0), the model predicts a log-unemployment of 1.402. Exponentiating this gives an estimated unemployment rate of 4.06%
# the R-Squared of 0.044 (adjusted R-Squared of 0.040) means the model explains just 4.4% of the variation in log unemployment
# With the F-statistic of 9.334, the overall model is statistically significant
```

6. Producing your Key Output

On your own, build up your model of logged unemployment rates gradually by also including (i) age and education variables; (ii) year as either a linear term or as a factor variable (you choose!) and finally (iii) country dummy variables to act as fixed effects picking up everything else that could differ across states. We won’t include lagged unemployment in our model as O & B do because of the risk of temporally auto correlated residuals (a breach of the independence assumption). 

When you’ve finished export (using stargazer) and briefly interpret your three models (i.e. the model with lagged homeowners, education and age; the model adding year and finally the model adding the country dummies). This table of model estimates plus your short interpretive text forms today’s Key Output for your technical appendix. What do the results tell us about the Oswald hypothesis? 

```{r}
# add lagged versions of 'Age65plus' and 'Degree' to the dataset
# the lag() function shifts each value down by one time period within each group
# thi sis useful for time series regression where you want to predict an outcome using prior-year information
europe_rates_long_merged <- europe_rates_long_merged %>% 
  mutate(Lag_age65plus = lag(Age65plus), # add a new column with previous year's % age 65+ for each country
         Lag_degree = lag(Degree) # add a new column with previous year's % with tertiary education for each country
         )
```

```{r}
# base model we previously tested
linear_model_1 <- lm(Log_unemp ~ Lag_own, data = europe_rates_long_merged)
stargazer(linear_model_1, type = "text")
```

```{r}
# base model + Lag_age65plus + Lag_degree
linear_model_2 <- lm(Log_unemp ~ Lag_own + Lag_age65plus + Lag_degree, data = europe_rates_long_merged)
stargazer(linear_model_2, type = "text")
# neither Lag_age65plus nor Lag_degree is statistically significant
# the addition of two variable led to an improvment in the R-Squared of only 0.001
# the lagged homeownership rate is the only significant predictor in the model
```

```{r}
# add Year as a control to account for time-specific effects
# using factor(Year) instead of a linear trends allows each year to have its own intercept
linear_model_3 <- lm(Log_unemp ~ Lag_own + Lag_age65plus + Lag_degree + factor(Year), data = europe_rates_long_merged)
stargazer(linear_model_3, type = "text")
# Lag_own has a statistically significant positive effect - a 1 percentage point increase in homeownership in the previous year is associated with a 1.2% increase in unemployment in the current year (after log transformtion)
# Lag_age65plus shows a significant positive effect - countrie swith more older residents tend to have slightly higher unemployment - possibly due to lower labour market turnover or generational job competition
# Lag_degree shows a marginally significant positive effect on unemployment, which may reflect mismatches between qualification and labour demand
# 2016, 2017, and 2018 show significantly negative coefficients, suggesting that unemployment rates decreased over time, independent of homeownership or demographics. This is likely due to post-crisis economic recovery across Europe
# the R-Squared of 0.198 shows that this model explains up to 20% of the variation in log unemployment
```

```{r}
# finally, as Country as a factor variable as a dummy variable to act as fixed effects picking up everything else that could differ across states
linear_model_4 <- lm(Log_unemp ~ Lag_own + Lag_age65plus + Lag_degree + factor(Year) + factor(Country), data = europe_rates_long_merged)
# inspect
stargazer(linear_model_4, type = "text")
# Lag_own's coefficient increase to 0.023, meaning that a 1% increase in lagged homeownership is associated with a 2.3% increase in the unemployment rate, holding other varialbes constant. This supports the Oswald hypothesis that higher homeownership may reduce labour mobility and increase unemployment
# Lag_age65plus - countries with a higher share of older adults in the previous year now statistically significantly reduce unemployment; more retirees in the population mean tighter labour supply?
# Lag_degree coefficient is small and not significant
# Reference year is 2012
# Years 2013-2016 show positive coefficients, suggesting unemployment was generally higher than in 2012, even controlling for other factors
# By 2018, the effect is close to zero, implying a return to 2012-level unemployment
# Reference country is Austria
# Countries like Greece (+1.63), Italy (+1.31), and Portugal (+0.88) have higher baseline unemployment than Austria, all else equal
# In contrast, Norway (-1.60), Ireland (-1.17), and Luxembourg (-1.41) have much lower baseline unemployment
# the model fit of 0.935 shows that thesse variable explain most of the variations in log unemployment rate
# The F-Statistic of 64.38 shows that the model as a whole is statistically significant
```

7. Evidence Brief 5 for the Policy Report

1. To what extent is a higher homeownership rate associated with increased youth unemployment across Europe?

The data on youth unemployment rates for our study period (from Eurostat) is available in a separate .csv file called 'Youth_unemployment_Europe.csv'. You'll need to read this into R, check it over and merge the information it contains onto the datasets you already have. Pay particular attention to missing values while doing this! Once you have your data prepared you then need to pick sensible methods to address the research question. You could use descriptives, regressions or a bit of both - it's entirely up to you.

Hint: youth here is defined as aged 16-24 and you don't need to look beyond the datasets I've given you for variables to answer this question. When producing your response, think about why we might expect youth unemployment to be particularly strongly or weakly affected by the homeownership rate. To do this you’ll need to consider the mechanisms Oswald argues could cause homeownership to raise unemployment  – are these likely to be particularly relevant for younger adults? Why or why not?

```{r}
# read in Youth_Unemployment_Europe.csv dataset
youth_unemployment <- read_csv(here("Youth_Unemployment_Europe.csv"))
```

```{r}
# merge the 'unemployment' and 'ownership' dataframes by the common column "Country"
europe_rates_youth <- merge(youth_unemployment, ownership, by="Country") %>% 
    mutate(across(starts_with("Youthunemp_"), as.numeric)) %>%
    filter(Country != "United Kingdom")
```

```{r}
# perform a Pearson correlation test between the 2011 youth unemployment rate and home ownership rate
cor.test(europe_rates_youth$Youthunemp_2018,
         europe_rates_youth$Own_2018,
         method = "pearson")

# correlation coefficient of 0.57 with a p-value of 0.01228 suggests a statistically significant, moderate positive relationship between youth unemployment and home ownership rates in 2011 across European countries
# the p-value (0.01228) is statistically significant at the 5% level (p < 0.05), meaning we can reject the null hypothesis that there is no correlation
```

```{r}
# ensure reproducibility by fixing the random seed. This helps geom_text_repel() place labels consistently across plots
set.seed(42)

# visualise the relationship between homeownership and youth  unemployment rates in 2011 across European conuntries
ggplot(data=europe_rates_youth, aes(Own_2011, Youthunemp_2011)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = 0.7) +
  geom_text_repel(aes(label=Country), size = 3.5, max.overlaps = 100) +
  labs(x = "Homeownership Rate in 2011 (%)", y = "Youth Unemployment Rate in 2011 (%)") +
  theme_classic() +
  theme(axis.text = element_text(size = 8))
  
# the regression line indicates a moderate positive correlation: countries with higher homeownership rates tend to have higher youth unemployment rates
# the shaded area represents the 95% confidence interval around the regression line
# Spain, Greece sit at the extreme upper-right, with both very high youth unemployment and high homeownerships, leading the trend
# Latvia, Croatia, Slovakia, and Lithuania also combine high homeownership with elevated youth unemployment, aligning with the overall pattern
# Switzerland, Austria, and Germany demonstrate the opposite pattern, with lower home ownership and low youth unemployment
# Romania and Norway are notable outliers in that they have very high homeownerhsip rates but comparatively low unemployment
# the plot supports the Oswald hypothesis that high homeownership may reduce labour mobility, potentially contributing to higher unemployment as young people may be less able or willing to relocate for work
# however, outliers like Norway or Romania suggest the relationship is not universal and may be hsaped by national policies, education systems, or broader economic factors
```

```{r}
# transform the dataset from wide to long format
europe_rates_youth_long <- europe_rates_youth %>% 
  pivot_longer(., cols = Youthunemp_2011:Own_2018, names_to = c(".value", "Year"), names_sep = "_")
```

```{r}
# calculate correlations by year and store in a data frame
cor_summary <- europe_rates_youth_long %>%
  group_by(Year) %>%
  summarise(
    cor = round(cor(Youthunemp, Own), 2),
    n = n()
  )

ggplot(cor_summary, aes(x = Year, y = cor)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(size = 2) +
  geom_text(aes(label = round(cor, 2)), vjust = -1.2, size = 3) +
  labs(
    # title = "Youth Unemployment–Homeownership Correlation Over Time",
    x = "Year",
    y = "Pearson's r"
  ) +
  ylim(0, 0.65) +
  theme_classic()


# output the correlation summary as a table using stargazer
stargazer(cor_summary, type = "text", summary = FALSE, rownames = FALSE)

# in 2011, the correlation between homeownership and youth unemployment was strongest at 0.57, indicating a moderate to strong positive relationship - countries with higher homeownership tended to have higher youth unemployment
# this correlation gradually declines over time - from 0.45 in 2012 to just 0.16 by 2018, suggesting that the association weakens as the European labour market recovers post-crisis
# this consistent downward trend may reflect:
# - economic recovery and improving labour condition across Europe
# - structural or policy changes that reduced the impact of homeownership on youth employment outcomes
# - a possible decoupling between housing tenure patterns and youth labour market performance
# by 2018, the correlation is quite weak (0.16), indicating that homeownership is no longer a strong predictor of youth unemployment
```

```{r}
# plot unemployment vs homeownership by year
ggplot(europe_rates_youth_long, aes(Own, Youthunemp)) +
  geom_point(colour = "navy", size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = 0.7) +
  labs(x = "Homeownership Rate (%)", y = "Youth unemployment Rate (%)") +
  theme_classic() +
  theme(axis.text = element_text(size = 8)) +
  facet_wrap(~Year)

# between 2011-2014, the relationship is clearly positive, supporting the hypothesis that high homeownership may hinder labour mobility
# the regression lines in these years have a noticeable upward slope and the correlation coefficients are moderate to strong
# in 2015 and 2016, the positive association starts to weaken, possibly due to post-crisis labour market recovery or housing policy adjustments
# the regression lines begin to flatten slightly and point clusters become more dispersed
# in 2017 and 2018, the relationship becomes very weak or almost flat, suggesting no clear association between homeownership and youth unemployment
# the diminishing correlation over time may reflect greater labour flexibility, improved job access, or changing housing dynamics
# overall, the initially strong positive relationship between homeownership and youth unemployment observed in the early 2010s dissipates over time
# this trend aligns with the idea that structural changes in labour markets and housing policy can disrupt previously observed patterns
```

```{r}
# calculate the change in unemployment and change in homeownership between 2011 and 2018
# the lage of 7 rows corresponds to a 7-year difference from 2011 to 2018
# keep only the 2018, so we're left with the change from 2011 to 2018 per country
changes <- europe_rates_youth_long %>%
  group_by(Country) %>%
  mutate(Change_youthunemp = Youthunemp - lag(Youthunemp, n =7), Change_own = Own - lag(Own, n = 7)) %>% 
  filter(Year == 2018) %>% 
  select(Country, Change_youthunemp, Change_own)

# inspect
changes
```

```{r}
# set seed
set.seed(42)

# vissualise how change in homeownership rates relate to changes in unemployment rates across European countries from 2011 to 2018
ggplot(data=changes, aes(Change_own, Change_youthunemp)) +
  geom_point() +
  geom_text_repel(aes(label=Country), size = 3.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "solid", linewidth = 0.7) +
  labs(x = "Change in Homeownership Rate 2011-2018", y = "Change in Youth Unemployment Rate 2011-2018") +
  theme_classic() +
  theme(axis.text = element_text(size = 8)) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_x_continuous(limits=c(-10, 5)) +
  scale_y_continuous(limits=c(-25, 5))

# Changes in homeownership rates are only weakly positively correlated with changes in youth unemployment rates between 2011 and 2018
# The regression line is nearly flat with a wide confidence interval, suggesting no strong or statistically significant relationship
# Most countries experienced declines in youth unemployment regardless of how homeownership rates changed
# Lithuania, Latvia, and Hungary show sharp declines in youth unemployment alongside stable or falling homeownership
# Ireland and Slovakia had substantial youth unemployment drops despite minimal homeownership changes
# Austria saw a modest increase in homeownership but little to no change in youth unemployment
# Denmark stands out with a significant decline in homeownership but only a modest decrease in youth unemployment
# Italy experienced a slight rise in youth unemployment with a minimal decline in homeownership
# Overall, this plot suggests that homeownership changes alone do not explain trends in youth unemployment over this period
```

```{r}
# fit a linear regression model to examine the relationship between changes in homeownership and changes in youth unemployment from 2011 to 2018
linear_1 <- lm(Change_youthunemp ~ Change_own, data = changes)

# display the summary of the regression model, including coefficients, R-squared value, and statistical significance
summary(linear_1)

# the summary results show that the relationship weakened consistenyl between 2011 and 2018
# on average, countries saw a decline of 7.40 percentage points in unemployment between 2011-2018 when there was no change in homeownerhsip
# slope: 0.26 (not significant, p = 0.62) - the relationship between changes in homeownership and changes in youth unemployment is weak and not statistically significant
# R-Squared of 0.009225 shows that less than 1% of the variation in unemployment change is explained by homeownership change. This is very weak explanatory power.
# do not include this in my analysis
```

```{r}
# transform the dataset from wide to long format
age_long <- pivot_longer(age, cols = Age65plus_2011:Age65plus_2018, names_to = "Year", names_prefix = "Age65plus_", values_to = "Age65plus")

# replace 'Slovak Republic' with 'Slovakia' to be consistent with other datasets
age_long <- age_long %>% 
  mutate(Country = ifelse(Country == "Slovak Republic", "Slovakia", Country))

# inspect
# Liechtenstein is missing from other datasets
# Croatia is missing from education_long dataset
age_long
```


```{r}
# transform the dataset from wide to long format
education_long <- pivot_longer(education, cols = Degree_2011:Degree_2018, names_to = "Year", names_prefix = "Degree_", values_to = "Degree")

# inspect
education_long
```

```{r}
# quck check for mismatched or missing countries
setdiff(age_long$Country, education_long$Country) # Liechtenstein missing from education_long
setdiff(education_long$Country, age_long$Country) # Croatia missing from age_long
```


```{r}
# merge age_long and education_long to our original europe_rates_long dataset
europe_rates_youth_long_merged <- europe_rates_youth_long %>% 
  merge(., age_long, by = c("Country", "Year")) %>% 
  merge(., education_long, by = c("Country", "Year"))
```

```{r}
# inspect - 224 rows remaining out of 232 (Croatia dropped. Liechtenstein was not in the original dataset. United Kingdom removed earlier)
nrow(europe_rates_youth_long_merged)
```

```{r}
# add a new column Log_unemp to noramlise exponential relationships
europe_rates_youth_long_merged <- mutate(europe_rates_youth_long_merged, Log_youthunemp = log(Youthunemp))
```

```{r}
# check that the Log_unemp variable has a more normal distribution than the raw Unemp rate variable
# histogram of raw unemployment rate
ggplot(europe_rates_youth_long_merged, aes(Youthunemp)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Youth Unemployment Rate", x = "Youth Unemployment Rate (%)", y = "Count") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# histogram of log-transformed unemployment rate
ggplot(europe_rates_youth_long_merged, aes(Log_youthunemp)) +
  geom_histogram(binwidth = 0.2, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Log Youth Unemployment Rate", x = "Log(Youth Unemployment Rate)", y = "Count") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

```{r}
# compute the lagged homeownership and unemployment rates for use in the model
# these variables just measure (i) homeownership and (ii) unemployment in the country in the previous year
europe_rates_youth_long_merged <- europe_rates_youth_long_merged %>%
  group_by(Country) %>%
  mutate(Lag_own = lag(Own, 1), Lag_log_youthunemp = lag(Log_youthunemp))
```

```{r}
# add lagged versions of 'Age65plus' and 'Degree' to the dataset
# the lag() function shifts each value down by one time period within each group
# this is useful for time series regression where you want to predict an outcome using prior-year information
europe_rates_youth_long_merged <- europe_rates_youth_long_merged %>% 
  mutate(Lag_age65plus = lag(Age65plus), # add a new column with previous year's % age 65+ for each country
         Lag_degree = lag(Degree) # add a new column with previous year's % with tertiary education for each country
         )
```

```{r}
# fit a linear model predicting log unemployment as a function just of lagged homeownership (this is our base model)
linear_model_1 <- lm(Log_youthunemp ~ Lag_own, data = europe_rates_youth_long_merged)

# use stargazer to present results
stargazer(linear_model_1, type = "text")

# for every 1 percentage point increase in lagged homeownership, the logged unemployment rate increase by 0.016 on average, holding everything else constant
# since log-unemployment is the outcome, this effect translate to a 1.6% increase in the unemployment rate per 1% increase in lagged homeownership
# the result is statistically significant at the 1% level
# at the baseline (when Lag_own = 0), the model predicts a log-unemployment of 1.699. Exponentiating this gives an estimated unemployment rate of 5.47%
# the R-Squared of 0.145 (adjusted R-Squared of 0.141) means the model explains 14.5% of the variation in log unemployment
# With the F-statistic of 33.895, the overall model is statistically significant
```

```{r}
# base model + Lag_age65plus + Lag_degree
linear_model_2 <- lm(Log_youthunemp ~ Lag_own + Lag_age65plus + Lag_degree, data = europe_rates_youth_long_merged)
stargazer(linear_model_2, type = "text")
# neither Lag_age65plus nor Lag_degree is statistically significant
# the addition of two variable led to an improvement in the R-Squared of only 0.008
# the adjusted R-Squared decreased in fact, from 0.141 to 0.139
# the lagged homeownership rate is the only significant predictor in the model
```

```{r}
# add Year as a control to account for time-specific effects
# using factor(Year) instead of a linear trends allows each year to have its own intercept
linear_model_3 <- lm(Log_youthunemp ~ Lag_own + Lag_age65plus + Lag_degree + factor(Year), data = europe_rates_youth_long_merged)
stargazer(linear_model_3, type = "text")
# Lag_own has a statistically significant positive effect - a 1 percentage point increase in homeownership in the previous year is associated with a 1.7% increase in youth unemployment in the current year (after log transformtion)
# Lag_age65plus shows a non-significant positive effect of 0.021 - countries with more older residents tend to have slightly higher unemployment - possibly due to lower labour market turnover or generational job competition
# Lag_degree shows a non-significant positive effect of 0.002 on youth-unemployment
# 2017 and 2018 show significantly negative coefficients, suggesting that unemployment rates decreased over time, independent of homeownership or demographics. This is likely due to post-crisis economic recovery across Europe
# the R-Squared of 0.250 shows that this model explains 25% of the variation in log unemployment, which is more than that for overall unemployment
```

```{r}
# finally, as Country as a factor variable as a dummy variable to act as fixed effects picking up everything else that could differ across states
linear_model_4 <- lm(Log_youthunemp ~ Lag_own + Lag_age65plus + Lag_degree + factor(Year) + factor(Country), data = europe_rates_youth_long_merged)
# inspect
stargazer(linear_model_4, type = "text")
# Lag_own's coefficient increased to 0.020, meaning that a 1% increase in lagged homeownership is associated with a 2% increase in the unemployment rate, holding other variables constant. This supports the Oswald hypothesis that higher homeownership may reduce labour mobility and increase unemployment
# Lag_age65plus - countries with a higher share of older adults in the previous year now statistically significantly reduce youth unemployment (-0.219); more retirees in the population mean tighter labour supply?
# Lag_degree coefficient is small and not significant
# Reference year is 2012
# Years 2013-2017 show positive coefficients, suggesting unemployment was generally higher than in 2012, even controlling for other factors
# By 2018, the effect is close to zero, implying a return to 2012-level unemployment
# Reference country is Austria
# Countries like Greece (+1.651), Italy (+1.621), and Portugal (+1.039) have higher baseline unemployment than Austria, all else equal
# In contrast, Norway (-0.948), Ireland (-0.822), and Luxembourg (-0.761) have much lower baseline unemployment
# the model fit of 0.95 shows that thesse variable explain most of the variations in log unemployment rate
# The F-Statistic of 84.531 shows that the model as a whole is statistically significant
```

```{r}
# finally, as we will take out the non-significant Lag_age65plus and Lag_degree variables so we can focus on the Lag_own variable while accounting for Year and Country factors.
linear_model_5 <- lm(Log_youthunemp ~ Lag_own + factor(Year) + factor(Country), data = europe_rates_youth_long_merged)
# inspect
stargazer(linear_model_5, type = "text")
# this led to a marginal decrease in the R-Squared to 0.940 (from 0.950)
# our Lag_own variable now is statistically significant at 99% level, with a coefficient of 0.027 - meaning for a 1% increases in the previous year's homeownership, there's a 2.7% increase in youth unemployment. This is a quite substantial effect.
# Reference year is 2012
# now, all the years show increasingly negative coefficients, with 2015-2018 statistically significant. This shows a consistently decreasing unemployment levels even after accounting for country-level variations
# Reference country is Austria
# Countries like Greece (+1.023), Italy (+0.781), and Portugal (+0.516) have higher baseline unemployment than Austria, all else equal
# In contrast, Norway (-0.670), Malta (-0.619) and Luxembourg (-0.761) have lower baseline unemployment
# the model fit of 0.95 shows that thesse variable explain most of the variations in log unemployment rate
# The F-Statistic of 74.147 shows that the model as a whole is statistically significant
```

